Environmental Conditions and Driver Behavior Affect Collision Outcomes in Toronto

### 2024-11-29 paper/paper.qmd
Refine the content of paper
Refine the paper and add some sections

---
title: "Environmental Conditions and Driver Behavior Shape Collision Outcomes in Toronto"
subtitle: "A Two-Stage Bayesian Analysis Reveals Infrastructure Factors Predict Crash Occurrence While Driver Actions Determine Severity"
author: Yitong Wang
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "This study employs a two stage bayesian modeling approach to analyze vehicle collision patterns in Toronto, examining how environmental conditions and driver behavior influence both injury occurrence and severity. The first model predicts the probability of injury based on environmental and infrastructural factors, while the second estimates injury severity using driver characteristics and behavior. Analysis of historical collision data from 2006 to 2023 reveals that while environmental conditions primarily influence injury occurrence, driver behavior factors like speeding and alcohol use more strongly affect severity levels. Our dual model analysis providing evidence based guidance for targeting both infrastructure improvements and enforcement strategies to reduce severe collision outcomes."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

# Introduction
Vehicle collisions in urban environments represent a significant public health and safety challenge, with particular concern in large metropolitan areas like Toronto where complex traffic patterns and diverse environmental conditions create varying levels of risk. Understanding the factors that contribute to collision occurrence and severity is crucial for urban planning and public safety policy, yet traditional analyses often treat these as a single outcome rather than recognizing their distinct underlying processes. The increasing availability of detailed collision data, combined with advances in statistical modeling, presents an opportunity to develop more nuanced understandings of how different factors influence collision outcomes.

@@

Our analysis reveals several distinct and important patterns in how different factors influence collision outcomes. Environmental conditions emerged as the dominant predictors of whether an injury occurs in the first place - particularly reduced visibility during winter months and poor road surface conditions on major arterials. However, once an injury occurs, driver behavior becomes the crucial determinant of severity. Impaired driving and speeding showed strong associations with severe outcomes, especially during nighttime hours and weekends. Road classification provided an interesting counterintuitive finding: while major arterial roads saw more frequent injuries, the injuries on local roads tended to be more severe. Temporal analysis uncovered clear risk patterns, with injury probability peaking during adverse weather conditions on arterials, while severity levels were highest during weekend nights with notable correlations to driver behavior violations. Pedestrian and cyclist collisions showed a particularly concerning pattern, even overall less frequent, they consistently resulted in more severe injuries across all conditions.

@@

The remainder of this paper is structured as follows:
@sec-data presents the overview of data (@sec-data-overview), data cleaning process (@sec-data-cleaning), measurement (@sec-data-measure), also for explanations, descriptions, graph summaries of outcome (@sec-data-outcome) and predictor variables (@sec-data-predictor) of the study. @sec-model set up the model and explain the process detailed, including procedure of model set up (@sec-model-setup) and model validation (@sec-model-validate), and also mention the assumption and validation (@sec-model-assumpandlimit) of models. In addition, (@sec-model-alternate) discuss the potential alternative models and its trade offs. Then, @sec-result reveal the prediction outcome and models' performance. @sec-discussion discusses the relation between the data and real world, which talks about the data quality and bias, as well as the potential limitation of our models in a broader context.

Appendix includes three parts: [Appendix -@sec-data-methodology] presents collision data methodology overview, [Appendix -@sec-data-idealized] provides detailed idealized methodology for collection improvement; [Appendix -@sec-model-details] presents additional summary and validation in modeling process.

@@

In this study, we utilized statistical language R [@citeR] and several key packages for data processing, analysis, and visualization. The tidyverse package [@tidyverse] facilitated data manipulation, with dplyr [@dplyr] being particularly crucial for data transformation and filtering operations, while visualization was conducted using ggplot2 [@ggplot2]. For efficient and consistent data import, we employed the readr package [@readr]. For handling temporal aspects of collision data, we employes lubridate {@lubridate} to process dates and times.We used here [@here] for reproducible file path management, arrow [@arrow] for efficient data storage and retrieval, and knitr [@knitr] used to provide the better formatting output. For unit testing and code validation, we employed the testthat package [@testthat] which provided a robust framework for testing our functions and ensuring code reliability. Brms package [@brms] plays the significant values for our model construction. 
For Bayesian analysis visualization, we employed bayesplot [@bayesplot] to create diagnostic plots including MCMC trace plots. For model evaluation, we employed the pROC package [@pROC] to generate receiver operating characteristic ROC curves and assess model performance. For model visualization and interpretation, we used the modelsummary package [@modelsummary] to create coefficient plots and model summaries.

## Data Cleaning {#sec-data-cleaning}
From the raw dataset, we selected 14 key variables most relevant to understanding collision severity patterns: date, time, road_class (road classification), accloc (accident location), traffctl (traffic control), visibility, light, rdsfcond (road surface condition), impactype (impact type), invage (involved person's age), injury (severity), speeding, ag_driv (aggressive driving), alcohol, and disability. Records with missing values in critical fields such as injury severity, visibility, or road classification were removed to ensure data quality.

In addition to removing rows with missing NA values, we also excluded records where key categorical variables were explicitly recorded as "None" despite an incident occurring. This includes records where accident location (accloc), road classification (road_class), traffic control (traffctl), visibility, light conditions (light), road surface condition (rdsfcond), and impact type (impactype) were recorded as "None". These "None" entries differ from missing values as they were actively recorded but indicate potentially incomplete or unclear documentation of the collision circumstances. This additional cleaning step ensures our analysis is based on collisions with well-documented characteristics, though it reduced our dataset by approximately 8% (1,057 records). This filtering helps maintain data quality and supports more reliable analysis of factor relationships with collision severity.

## Measurement {#sec-data-measure}
The transformation of real-world collision incidents into structured data points follows a rigorous protocol established by Toronto Police Services. When a collision occurs, first responders and investigating officers collect data through a standardized Motor Vehicle Accident Report Form. This form ensures systematic documentation of collision characteristics, converting complex real-world events into quantifiable measurements. For instance, injury severity classification follows specific criteria: fatalities are recorded when death occurs within 30 days of collision, major injuries are those requiring hospitalization, minor injuries need medical attention but no hospitalization, and minimal injuries indicate no significant medical intervention was required. This standardized approach helps maintain consistency in severity classification across different incidents and responding officers [@toronto_police_services_ksi_guide].

@@

Some driver related factors present particular measurement challenges due to their complex nature. Alcohol involvement is measured through standardized tests when possible, and it providing more objective data. However, other driver conditions like fatigue or distraction often rely on post incident interviews and officer's observation, which introducing potential reporting biases. Additionally, factors like driver age and license status are verified through official documentation, ensuring accuracy in demographic data. The measurement process also captures infrastructure characteristics such as road classification and traffic control presence, which are determined through existing municipal records and on-site verification. This multi-faceted approach to data collection, while comprehensive, means that the quality and reliability of different variables may vary depending on the objectivity of their measurement methods.

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-outcome
#| fig-cap: "Outcome frequency"
cleaned_data %>%
  ggplot(aes(x = injury, fill = injury)) +
  geom_bar() +
  scale_fill_brewer(palette = "RdYlBu") +
  labs(x = "Injury Severity",
       y = "Number of Collisions") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::comma)
```
@fig-outcome The primary outcome variable in our analysis is collision severity, represented through an ordered factor ranging from "None" to "Fatal". It illustrates the distribution of injury severity across all recorded incidents. The data reveals an interesting bimodal distribution in collision outcomes, with concentrations at both ends of the severity spectrum. Major injuries and no injuries represent the most frequent outcomes, accounting for 40.1% and 38.9% of incidents respectively. Fatal collisions comprise 6.3% of the dataset, while minor and minimal injuries together represent approximately 15% of cases.

@@

@fig-visibility Visibility factor plays a crucial role in collision outcomes. Figure 2 presents the relationship between visibility conditions and collision severity. Clear conditions account for the majority of recorded incidents (86.4%), while poor visibility conditions (including rain, snow, and freezing rain) represent 10.5% of cases. Notably, the proportion of severe injuries increases under poor visibility conditions, suggesting a significant relationship between environmental conditions and collision outcomes.

@@

@fig-road Road classification and traffic control measures represent important infrastructure factors in our analysis. Figure 3 illustrates the distribution of collisions across different road types and their associated severity levels. Major arterial roads account for 71.3% of all collisions, followed by minor arterial roads at 15.7%. This distribution likely reflects both traffic volume and speed limits on these road types. The relationship between road classification and injury severity reveals important patterns for urban safety planning.

@@

# Create a facet grid visualization with RdYlBu color palette
ggplot(interaction_data, aes(x = light, y = proportion, fill = injury)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ visibility, scales = "free_y") +
  scale_fill_brewer(palette = "RdYlBu") +
  labs(
    title = "Interaction Between Visibility and Light on Injury Severity",
    x = "Light Conditions",
    y = "Proportion of Collisions",
    fill = "Injury Severity"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@fig-visibilityandlight The facet grid visualization reveals the interaction between visibility and light conditions on injury severity. Under poor visibility (fog), dark conditions show a noticeably higher proportion of severe injuries (Fatal, Major) compared to daylight, indicating the compounded risk of low visibility and inadequate lighting. In contrast, clear visibility with daylight consistently shows the lowest proportions of severe injuries, emphasizing the protective role of good environmental conditions. These findings will help guide the inclusion of visibility and light in the predictive model, ensuring it accounts for the impact of environmental factors on injury outcomes. This approach will improve the model's ability to identify scenarios with higher risks of severe injuries.

@@

# Create line chart for proportions
ggplot(proportional_data, aes(x = injury, y = proportion, group = status, color = status)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  facet_wrap(~ behavior, scales = "free_y") +
  labs(
    title = "Proportion of Injury Severity by Driver Behaviors (Yes/No)",
    x = "Injury Severity",
    y = "Proportion of Collisions",
    color = "Behavior Status"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```
@fig-behaviors The proportional line chart compares injury severity distributions for drivers involved in risky behaviors (speeding, alcohol use, aggressive driving) versus those who were not. It reveals that drivers engaged in these behaviors often have higher proportions of severe injuries, such as Fatal or Major, compared to those who did not. However, the results may be influenced by sample size imbalances, as there are likely fewer drivers in the "Yes" group for alcohol or aggressive driving, which could exaggerate proportions and introduce bias. Despite these limitations, the chart provides valuable insights into the heightened risks associated with these behaviors

@@

@fig-impacttype The stacked bar chart reveals the proportion of injury severities across the top five collision impact types, providing insight into how different types of collisions relate to injury outcomes. For instance, pedestrian collisions show a significantly higher proportion of severe injuries (Fatal and Major), while rear-end collisions are dominated by minor or no injuries. This exploration is critical for our paper as it identifies which collision types pose the greatest risk of severe outcomes, helping us understand the underlying patterns of injury severity. By focusing on this relationship, we address a core question of our study: how specific predictors, such as impact type, influence injury severity.

@@

# Model {#sec-model}
This study employs a two-stage modeling approach to understand and predict traffic injury outcomes, a model for injury occurrence prediction and another model for injury severity prediction when injuries occur. Both models were implemented using the brms package [@brms] in R, leveraging its robust framework for Bayesian regression modeling. The dual-model Bayesian framework precisely analyze traffic injuries, separating the prediction of injury occurrence from injury severity assessment. This approach allows us to isolate and examine the distinct factors that influence whether an injury occurs and, subsequently, how severe that injury becomes. By developing separate models for these two aspects, we can better understand the complex dynamics of traffic accidents and provide more targeted insights for safety interventions.

@@

This model addresses the binary outcome of injury occurrence through a Bayesian logistic regression framework. Let $Y_i$ represent the injury outcome for observation $i$, where $Y_i = 1$ indicates an injury occurred and $Y_i = 0$ indicates no injury. The probability of injury is modeled through a Bernoulli distribution:

@@

The occurrence model prioritizes environmental and infrastructural factors, drawing on established traffic safety literature and the patterns observed in our exploratory data analysis. Environmental conditions are represented through visibility, light conditions, and road surface state, each categorized to capture fundamental differences while maintaining model parsimony. We consolidate visibility into a binary indicator (Good/Poor) based on detailed weather conditions, simplify lighting into day/night categories from six original conditions, and classify road conditions as good or poor based on surface characteristics.
Infrastructure factors incorporate the physical and operational aspects of the road environment. Road classification follows a hierarchical structure (Major/Minor/Local) that reflects traffic volume and design standards. Location type captures the crucial distinction between intersections, private drives, and other locations, while traffic control variables account for the presence and type of traffic management systems. Additionally, temporal factors such as rush hour periods and weekend indicators are included to capture systematic variations in traffic patterns.

### Injury severity model
Our severity model employs a more nuanced approach, focusing exclusively on collisions where injuries occurred. We first excluded all cases classified as "None" injury, as these fall outside the scope of severity analysis. Among injury cases, we created a binary classification that distinguishes between severe and minor injuries. We classified "Fatal" and "Major" injuries as "Severe" outcomes, while grouping "Minimal" and "Minor" injuries as "Minor" outcomes.

For cases where injuries occur, our second model examines severity levels using an ordered logistic regression approach. The severity outcome $S_i$ is modeled as:

@@

The severity model emphasizes driver behavioral and impact factors that influence injury outcomes once an accident occurs. Driver behavior variables include binary indicators for speeding, aggressive driving, alcohol involvement, and driver disability. Age groups are categorized to capture the varying vulnerability and risk patterns across different life stages, while maintaining sufficient group sizes for reliable estimation.
Impact characteristics play a crucial role in determining injury severity. The model incorporates impact type as a categorical variable distinguishing between pedestrian, cyclist, rear-end, and angle collisions, each representing distinct injury risk patterns. Environmental conditions are represented through light conditions and road classification, which can moderate the relationship between impact forces and resulting injuries.

## Model validation {#sec-model-validate}
Our validation approach employs a sophisticated stratified sampling methodology, dividing the data into training (70%) and testing (30%) sets while carefully maintaining the original class proportions. This strategy is particularly crucial given the inherently imbalanced nature of traffic injury data. The validation framework encompasses multiple complementary approaches, including confusion matrix analysis for understanding classification performance, receiver operating characteristic (ROC) curve analysis for assessing discriminative ability. We further ensure model stability through comprehensive cross validation procedures. Prior to model fitting, we conducted thorough multicollinearity assessments using Variance Inflation Factors, confirming the independence of our predictor variables.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-posterior
#| fig-cap: "Posterior Predictive Check for Injury Occurrence Model"
pp_check(occurrence_model, type = "bars", ndraws = 100) +
  ggtitle("Posterior Predictive Check for Injury Occurrence Model")
```
@fig-posterior The posterior predictive check plot demonstrates the model's ability to replicate the binary nature of injury occurrence data. The blue dots (yrep) represent simulated data from the posterior predictive distribution, while the light blue bars (y) show the actual observed data distribution. The plot reveals two clear peaks around 0 and 1, corresponding to no-injury and injury cases respectively, with the simulated data closely matching the observed distribution. The alignment between predicted and actual counts at both peaks suggests that the model effectively captures the underlying binary structure of injury occurrences while maintaining appropriate uncertainty in its predictions.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-trace_o1
#| fig-cap: "Trace Plots for first 7 parameters in Injury Occurrence Model"
# First 7 parameters
mcmc_trace(as_draws(occurrence_model), 
          regex_pars = "b_(Intercept|visibilityPoor|lightNight|rdsfcondPoor|road_classMajor|road_classMinor|acclocOther)") +
          ggtitle("Trace Plots for Injury Occurrence Model - Part 1")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-trace_o2
#| fig-cap: "Trace Plots for middle 7 parameters in Injury Occurrence Model"
# Next 7 parameters
mcmc_trace(as_draws(occurrence_model), 
          regex_pars = "b_(acclocPrivate|traffctlOther|traffctlSignal|traffctlStop|impactypeApproaching|impactypeCyclistCollisions|impactypeOther)") +
          ggtitle("Trace Plots for Injury Occurrence Model - Part 2")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-trace_o3
#| fig-cap: "Trace Plots for last 6 parameters in Injury Occurrence Model"
# Last 8 parameters
mcmc_trace(as_draws(occurrence_model), 
          regex_pars = "b_(impactypePedestrianCollisions|impactypeRearEnd|impactypeSideswipe|impactypeSMVOther|impactypeSMVUnattendedVehicle|impactypeTurningMovement)") +
          ggtitle("Trace Plots for Injury Occurrence Model - Part 3")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-trace_s1
#| fig-cap: "Trace Plots for first six parameters in Injury Secerity Model"
# First plot - first group of parameters
mcmc_trace(as_draws(severity_model), 
          regex_pars = "b_(Intercept|speeding|ag_driv|alcohol|disability|age)") +
          ggtitle("Trace Plots for Injury Severity Model - Part 1")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-trace_s2
#| fig-cap: "Trace Plots for rest parameters in Injury Secerity Model"
# Second plot - second group of parameters
mcmc_trace(as_draws(severity_model), 
          regex_pars = "b_(impact|light|road)") +
          ggtitle("Trace Plots for Injury Severity Model - Part 2")
```
@fig-trace_s1, @fig-trace_s2 The trace plots for various model parameters display stable MCMC sampling convergence across all four chains (represented in different shades of blue). Each parameter's trace shows consistent exploration within its respective range without any concerning patterns or trends, indicating good mixing of the chains. The overlapping nature of the chains and their stationary behavior throughout the sampling period provides strong evidence that the model has reached convergence and the posterior distributions are reliable estimates of the true parameter values.

@@

# Results {#sec-result}
## Model result
Our dual-model analysis of traffic injuries reveals distinct patterns in both injury occurrence and severity prediction, with both models demonstrating strong predictive performance across different scenarios and conditions. Key findings from our models reveal the relative importance of different predictors in determining both injury occurrence and severity, and we summarise it in the follow coefficients tables and ROC curve plot.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: tbl-occurrence
#| tbl-cap: "Occurrence Model Coefficients"
# For occurrence model
occurrence_summary <- as.data.frame(summary(occurrence_model)$fixed)
occurrence_table <- occurrence_summary %>%
  rownames_to_column("Variable") %>%
  select(Variable, Estimate, `Est.Error`, `l-95% CI`, `u-95% CI`, `Bulk_ESS`, `Tail_ESS`) %>%
  rename(
    'Std. Error' = 'Est.Error',
    'Lower CI' = 'l-95% CI',
    'Upper CI' = 'u-95% CI',
    'Bulk ESS' = 'Bulk_ESS',
    'Tail ESS' = 'Tail_ESS'
  )

# Create formatted tables
kable(occurrence_table,
      format = "pipe",
      digits = 3)
```

>>>>

---
title: "Environmental Conditions and Driver Behavior Affect Collision Outcomes in Toronto"
subtitle: "A Two-Stage Bayesian Analysis Shows Infrastructure Factors Predict Crash Occurrence While Driver Actions Determine Severity"
author: Yitong Wang
thanks: "Code and data are available at: [https://github.com/demainwang/Toronto_vehicle_collision_analysis](https://github.com/demainwang/Toronto_vehicle_collision_analysis)."
date: today
date-format: long
abstract: "This study employs a two stage bayesian modeling approach to analyze vehicle collision patterns in Toronto, examining how environmental conditions and driver behavior influence both injury occurrence and severity. The first model predicts the probability of injury based on environmental and infrastructural factors, while the second estimates injury severity using driver characteristics and behavior. Analysis of historical collision data from 2006 to 2023 shows that while environmental conditions primarily influence injury occurrence, driver behavior factors like speeding and alcohol use more strongly affect severity levels. These findings demonstrate the need for a dual-pronged approach to urban safety through targeted infrastructure improvements in high-risk areas to prevent collisions, alongside enhanced enforcement strategies focusing on dangerous driving behaviors. "
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

# Introduction
Vehicle collisions in urban environments represent a significant public health and safety challenge, with particular concern in large metropolitan areas like Toronto where complex traffic patterns and diverse environmental conditions create varying levels of risk. Understanding the factors that contribute to collision occurrence and severity is important for urban planning and public safety policy, yet traditional analyses often treat these as a single outcome rather than recognizing their distinct underlying processes. The increasing availability of detailed collision data, combined with advances in statistical modeling, presents an opportunity to develop more precise understandings of how different factors influence collision outcomes.

@@

Our analysis shows several distinct and important patterns in how different factors influence collision outcomes. Environmental conditions emerged as the dominant predictors of whether an injury occurs in the first place - particularly reduced visibility during winter months and poor road surface conditions on major arterials. However, once an injury occurs, driver behavior becomes a determinant of severity. Impaired driving and speeding showed strong associations with severe outcomes, especially during nighttime hours and weekends. Road classification provided an interesting counterintuitive finding: while major arterial roads saw more frequent injuries, the injuries on local roads tended to be more severe. Temporal analysis uncovered clear risk patterns, with injury probability peaking during adverse weather conditions on arterials, while severity levels were highest during weekend nights with notable correlations to driver behavior violations. Pedestrian and cyclist collisions showed a particularly concerning pattern, even overall less frequent, they consistently resulted in more severe injuries across all conditions.

@@

The remainder of this paper is structured as follows:
@sec-data presents the overview of data (@sec-data-overview), data cleaning process (@sec-data-cleaning), measurement (@sec-data-measure), also for explanations, descriptions, graph summaries of outcome (@sec-data-outcome) and predictor variables (@sec-data-predictor) of the study. @sec-model set up the model and explain the process detailed, including procedure of model set up (@sec-model-setup) ,and also mention the assumption and validation (@sec-model-assumpandlimit) of models. In addition, (@sec-model-alternate) discuss the potential alternative models and its trade offs. Then, @sec-result presents the prediction outcome and models' performance. @sec-discussion discusses the relation between the data and real world, which talks about the data quality and bias, as well as the potential limitation of our models in a broader context.

Appendix includes three parts: [Appendix -@sec-data-methodology] presents collision data methodology overview, [Appendix -@sec-data-idealized] provides detailed idealized methodology for collection improvement; [Appendix -@sec-model-validate_and_details] presents model validation (@sec-model-validate) and additional details in modeling process (@sec-model-detail).

@@

In this study, we utilized statistical language R [@citeR] and several key packages for data processing, analysis, and visualization. The tidyverse package [@tidyverse] facilitated data manipulation, with dplyr [@dplyr] for data transformation and filtering operations, while visualization was conducted using ggplot2 [@ggplot2]. For efficient and consistent data import, we employed the readr package [@readr]. For handling temporal aspects of collision data, we employes lubridate (@lubridate) to process dates and times.We used here [@here] for reproducible file path management, arrow [@arrow] for efficient data storage and retrieval, and knitr [@knitr] used to provide the better formatting output. For unit testing and code validation, we employed the testthat package [@testthat] which provided a robust framework for testing our functions and ensuring code reliability. Brms package [@brms] plays the significant values for our model construction. 
For Bayesian analysis visualization, we employed bayesplot [@bayesplot] to create diagnostic plots including MCMC trace plots. For model evaluation, we employed the pROC package [@pROC] to generate receiver operating characteristic ROC curves and assess model performance. For model visualization and interpretation, we used the modelsummary package [@modelsummary] to create coefficient plots and model summaries.

## Data Cleaning {#sec-data-cleaning}
From the raw dataset, we selected 14 key variables most relevant to understanding collision severity patterns: date, time, road_class (road classification), accloc (accident location), traffctl (traffic control), visibility, light, rdsfcond (road surface condition), impactype (impact type), invage (involved person's age), injury (severity), speeding, ag_driv (aggressive driving), alcohol, and disability. Records with missing values in important fields such as injury severity, visibility, or road classification were removed to ensure data quality.

In addition to removing rows with missing NA values, we also excluded records where key categorical variables were explicitly recorded as "None" despite an incident occurring. This includes records where accident location (accloc), road classification (road_class), traffic control (traffctl), visibility, light conditions (light), road surface condition (rdsfcond), and impact type (impactype) were recorded as "None". These "None" entries differ from missing values as they were actively recorded but indicate potentially incomplete or unclear documentation of the collision circumstances. This additional cleaning step ensures our analysis is based on collisions with well-documented characteristics, though it reduced our dataset by approximately 8% (1,057 records).

## Measurement {#sec-data-measure}
The transformation of real-world collision incidents into structured data follows a rigorous protocol established by Toronto Police Services. When a collision occurs, first responders and investigating officers collect data through a standardized Motor Vehicle Accident Report Form. This form ensures systematic documentation of collision characteristics, converting complex real-world events into quantifiable measurements. For instance, injury severity classification follows specific criteria: fatalities are recorded when death occurs within 30 days of collision, major injuries are those requiring hospitalization, minor injuries need medical attention but no hospitalization, and minimal injuries indicate no significant medical intervention was required. This standardized approach helps maintain consistency in severity classification across different incidents and responding officers [@toronto_police_services_ksi_guide].

@@

Some driver related factors present particular measurement challenges due to their complex nature. Alcohol involvement is measured through standardized tests if it is possible, and it providing more objective data. However, other driver conditions like fatigue or distraction often rely on post incident interviews and officer's observation, which introducing potential reporting biases. Additionally, factors like driver age and license status are verified through official documentation, ensuring accuracy in demographic data. The measurement process also captures infrastructure characteristics such as road classification and traffic control presence, which are determined through existing municipal records and on site verification. Even this multi faceted approach makes data is well covered, it also means the quality and reliability of different variables may vary depending on the objectivity of their measurement methods.

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-outcome
#| fig-cap: "Outcome frequency"
cleaned_data %>%
  ggplot(aes(x = injury, fill = injury)) +
  geom_bar() +
  scale_fill_brewer(palette = "RdYlBu") +
  labs(x = "Injury Severity",
       y = "Number of Collisions") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::comma)
```
@fig-outcome The primary outcome variable in our analysis is collision severity, represented through an ordered factor ranging from "None" to "Fatal". It illustrates the distribution of injury severity across all recorded incidents. The data shows an interesting bimodal distribution in collision outcomes, with concentrations at both ends of the severity spectrum. Major injuries and no injuries represent the most frequent outcomes, accounting for 40.1% and 38.9% of incidents respectively. Fatal collisions comprise 6.3% of the dataset, while minor and minimal injuries together represent approximately 15% of cases.

@@

@fig-visibility Visibility factor plays a important role in collision outcomes. It presents the relationship between visibility conditions and collision severity. Clear conditions account for the majority of recorded incidents (86.4%), while poor visibility conditions (including rain, snow, and freezing rain) represent 10.5% of cases. Notably, the proportion of severe injuries increases under poor visibility conditions, suggesting a significant relationship between environmental conditions and collision outcomes.

@@

@fig-road illustrates the distribution of collisions across different road types and their associated severity levels. Major arterial roads account for 71.3% of all collisions, followed by minor arterial roads at 15.7%. This distribution likely reflects both traffic volume and speed limits on these road types.

@@

# Create a facet grid visualization with RdYlBu color palette
ggplot(interaction_data, aes(x = light, y = proportion, fill = injury)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ visibility, scales = "free_y") +
  scale_fill_brewer(palette = "RdYlBu") +
  labs(
    title = "Interaction Between Visibility and Light on Injury Severity",
    x = "Light Conditions",
    y = "Proportion of Collisions",
    fill = "Injury Severity"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1))
```

@fig-visibilityandlight The facet grid visualization presents the interaction between visibility and light conditions on injury severity. Under poor visibility (fog), dark conditions show a noticeably higher proportion of severe injuries (Fatal, Major) compared to daylight, indicating the compounded risk of low visibility and inadequate lighting. In contrast, clear visibility with daylight consistently shows the lowest proportions of severe injuries, emphasizing the protective role of good environmental conditions. These findings will help guide the inclusion of visibility and light in the predictive model, ensuring it accounts for the impact of environmental factors on injury outcomes.

@@

# Create line chart for proportions
ggplot(proportional_data, aes(x = injury, y = proportion, group = status, color = status)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  facet_wrap(~ behavior, scales = "free_y") +
  labs(
    title = "Proportion of Injury Severity by Driver Behaviors (Yes/No)",
    x = "Injury Severity",
    y = "Proportion of Collisions",
    color = "Behavior Status"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```
@fig-behaviors The proportional line chart compares injury severity distributions for drivers involved in risky behaviors (speeding, alcohol use, aggressive driving) versus those who were not. It shows that drivers engaged in these behaviors often have higher proportions of severe injuries, such as Fatal or Major, compared to those who did not. However, the results may be influenced by sample size imbalances, as there are likely fewer drivers in the "Yes" group for alcohol or aggressive driving, which could exaggerate proportions and introduce bias.

@@

@fig-impacttype The stacked bar chart demonstrates the proportion of injury severities across the top five collision impact types, providing understand for how different types of collisions relate to injury outcomes. For instance, pedestrian collisions show a significantly higher proportion of severe injuries (Fatal and Major), while rear-end collisions are dominated by minor or no injuries.

@@

# Model {#sec-model}
This study employs a two-stage modeling approach to understand and predict traffic injury outcomes, a model for injury occurrence prediction and another model for injury severity prediction when injuries occur. Both models were implemented using the brms package [@brms] in R, the dual-model Bayesian framework precisely analyze traffic injuries, separating the prediction of injury occurrence from injury severity assessment. This approach allows us to isolate and examine the distinct factors that influence whether an injury occurs and, subsequently, how severe that injury becomes. By developing separate models for these two aspects, we can better understand the complex dynamics of traffic accidents.

@@

Let $Y_i$ represent the injury outcome for observation $i$, where $Y_i = 1$ indicates an injury occurred and $Y_i = 0$ indicates no injury. The probability of injury is modeled through a Bernoulli distribution:

@@

The occurrence model prioritizes environmental and infrastructural factors, drawing on established traffic safety literature and the patterns observed in our exploratory data analysis. Environmental conditions are represented through visibility, light conditions, and road surface state, each categorized to capture fundamental differences while maintaining model parsimony. We consolidate visibility into a binary indicator (Good/Poor) based on detailed weather conditions, simplify lighting into day/night categories from six original conditions, and classify road conditions as good or poor based on surface characteristics.
Infrastructure factors incorporate the physical and operational aspects of the road environment. Road classification follows a hierarchical structure (Major/Minor/Local) that reflects traffic volume and design standards. Location type captures the distinction between intersections, private drives, and other locations, while traffic control variables account for the presence and type of traffic management systems. Additionally, temporal factors such as rush hour periods and weekend indicators are included to capture systematic variations in traffic patterns.

### Injury severity model
Our severity model is focusing exclusively on collisions where injuries occurred. We first excluded all cases classified as "None" injury, as these fall outside the scope of severity analysis. Among injury cases, we created a binary classification that distinguishes between severe and minor injuries. We classified "Fatal" and "Major" injuries as "Severe" outcomes, while grouping "Minimal" and "Minor" injuries as "Minor" outcomes.

For cases where injuries occur, the severity outcome $S_i$ is modeled as:

@@

The severity model emphasizes driver behavioral and impact factors that influence injury outcomes once an accident occurs. Driver behavior variables include binary indicators for speeding, aggressive driving, alcohol involvement, and driver disability. Age groups are categorized to capture the varying vulnerability and risk patterns across different life stages, while maintaining sufficient group sizes for reliable estimation.
Impact characteristics play a imporatant role in determining injury severity. The model incorporates impact type as a categorical variable distinguishing between pedestrian, cyclist, rear-end, and angle collisions, each representing distinct injury risk patterns. Environmental conditions are represented through light conditions and road classification, which can moderate the relationship between impact forces and resulting injuries.

@@

# Results {#sec-result}
## Model result
Our dual model analysis of traffic injuries demonstrates distinct patterns in both injury occurrence and severity prediction, key findings from our models show the relative importance of different predictors in determining both injury occurrence and severity, and we summarise it in the follow tables and ROC curve plot.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: tbl-summary_modeloccurrence
#| tbl-cap: "Summary for Occurrence model"
modelsummary(
  occurrence_model,
  output = "kableExtra",
) 
```

@tbl-summary_modeloccurrence The Occurrence Model uncovers several noteworthy patterns in predicting the likelihood of injuries in traffic collisions. Among the environmental factors, nighttime conditions (0.131) are associated with a higher likelihood of injury, highlighting the risks associated with reduced visibility and nighttime driving.
Impact types emerge as the most influential predictors. Single motor vehicle (SMV) incidents show the strongest positive association with injury occurrence (1.224), likely reflecting the dynamics of these collisions, such as loss of control or impacts with stationary objects. Interestingly, cyclist (-0.773) and pedestrian (-0.772) collisions exhibit substantial negative associations with injury occurrence. This might seem contradictory to the findings in our Severity Model, where these types of collisions are strongly linked to severe injuries. However, the explanation may lie in the polarizing nature of these incidents many result in no injury at all or extreme injuries, which could skew their average likelihood of injury occurrence.

Most infrastructure-related variables, such as road classification and traffic control measures, show relatively modest or non-significant effects, with confidence intervals crossing zero. This suggests that their influence on injury occurrence might depend heavily on specific contexts or interactions with other factors.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: tbl-summary_modelseverity
#| tbl-cap: "Summary for Severity model"
modelsummary(
  severity_model,
  output = "kableExtra",
) 
```

@tbl-summary_modelseverity The model analysis is based on 4,572 cases that injury happened in collision, with an R-squared value of 0.111 indicating that about 11.1% of the variance in injury severity is explained by the included predictors. Collisions involving cyclists (2.31) and pedestrians (1.73) stand out as the most significant predictors of severe outcomes, emphasizing the heightened vulnerability of these road users. Senior individuals are also more likely to experience severe injuries, reflecting the increased physical fragility of this age group. Interestingly, collisions on minor roads are associated with a higher likelihood of severe injuries, possibly due to less stringent traffic control measures or more complex road layouts. On the other hand, factors like speeding and alcohol involvement, while often associated with risky driving behaviors, did not show a strong connection to injury severity in this dataset.
